---
title: "Data Science HW 2"
author: Hingling Yu
date: 10/01/2023
output: github_document
geometry: margin=0.85in
fontsize: 12pt
  
---

## Problem 0

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, message = FALSE)
```

```{r load_libraries}
library(tidyverse)
library(readxl)
library(janitor)
library(dplyr)
```

## Problem 1

### First Step: Cleaning `pols-month.csv`

```{r clean_538_pols}
pols = 
  read_csv("data/data_fivethirtyeight/pols-month.csv") %>% #import the `pols-month.csv` file
  separate(mon, into = c("year", "month", "day"), sep = "-") %>% # split the mon column into three columns: year, month, day using "="
  mutate(
    month = month.name[as.numeric(month)], # converts numeric month values to full month names
    president = ifelse(prez_gop == 1, "gop", "dem"), # creates a new variable president, assign 'gop' if prez_gop = 1, otherwise assign 'dem'
    year = as.numeric(year)
  ) %>% 
  select(year, month, everything(), -day, -starts_with("prez")) # select all variables except 'day' and everything starts with 'prez'
```



### Second Step: Cleaning `snp.csv`

```{r clean_538_snp}
snp = 
  read_csv("data/data_fivethirtyeight/snp.csv") %>% #import the `snp.csv` file
  separate(date, into = c("month", "day", "year"), sep = "/") %>% # split the data column into three columns: month, day, year using "/"
  mutate(
    month = month.name[as.numeric(month)], 
    year = as.numeric(year),
    year = ifelse(year < 50, year + 2000, year + 1900) # add 2000 to years value less than 50, 1900 to years value greater or equal to 50
  ) %>% #converts numeric month to full month names
  select(year, month, close, -day) # rearrange the columns so "year" and "month" at first, then "close". Remove "day"


```


### Third Step: Tidy `umeployment.csv`

```{r clean_538_unemployment}
unemployment = 
  read_csv("data/data_fivethirtyeight/unemployment.csv") %>%  #import the `unemployment.csv`
  rename(year = Year) %>% # rename 'Year' column to 'year'
  pivot_longer(
    Jan:Dec, # convert from wide to long
    names_to = "month", # names of the columns into a new column named 'month'
    values_to = "unemployment_rate"# values of the columns into a new column named 'unemployment_rate'
    ) %>% 
  mutate(
    month = month.name[match(month, month.abb)],
    year = as.numeric(year)
    ) %>% # finds the position of each month abbreviation name in month.abb vector then concert these month abbreviation names to full month names
  select(year, month, unemployment_rate) 

```


### Last Step: Merge three datasets


```{r merge_538}
data_538 =
  left_join(pols, snp) %>% 
  left_join(unemployment) 

str(data_538)
```



## Problem 2

### Clean Mr.Trash Wheel


```{r clean_mr_trash_wheel}
mr_trash_wheel =
  read_excel("data/data_trashwheel/trash_wheel_collection.xlsx", sheet = 1) %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>% 
  select(-x15,-x16) %>% 
  mutate(
    homes_powered = (weight_tons * 500) / 30,
    wheel_name = "Mr. Trash Wheel",
    year = as.numeric(year)
  )

```



### Clean Professor. Trash Wheel

```{r clean_professor_trash_wheel}
prof_trash_wheel =
  read_excel("data/data_trashwheel/trash_wheel_collection.xlsx",sheet = 2) %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>% 
  mutate(
    homes_powered = (weight_tons * 500) / 30,
    wheel_name = "Professor Trash Wheel",
    year = as.numeric(year)
  )
  
  
```



### Clean Gwynnda Trash Wheel

```{r clean_gwynnda_trash_wheel}
gwy_trash_wheel = 
  read_excel("data/data_trashwheel/trash_wheel_collection.xlsx", sheet = 4) %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster) %>% 
  mutate(
    homes_powered = (weight_tons * 500) / 30,
    wheel_name = "Gwynnda Trash Wheel",
    year = as.numeric(year)
  )

```


### Combine all `mr_trash_wheel`, `prof_trash_wheel`, and `gwy_trash_wheel`

```{r merge_trash_wheel_combine}
trash_wheel_combine = 
  bind_rows(mr_trash_wheel, prof_trash_wheel, gwy_trash_wheel)

str(trash_wheel_combine)

```


## Description of Problem 2

- For the `mr_trash_wheel` dataset, there are `r nrow(mr_trash_wheel)` observations and `r ncol(mr_trash_wheel)` variables. This data provides detail records of waste collection, capturing various information of waste items collected, ranging from years `r mr_trash_wheel |> pull(year) |> min()` to `r mr_trash_wheel |> pull(year) |> max()`. The key variables includes `dumpster`, `weight_tons`, `homes_powered`. 

-  For the `prof_trash_wheel` dataset, there are `r nrow(prof_trash_wheel)` observations and `r ncol(prof_trash_wheel)` variables. This data also provides detail records of waste collection, capturing various information of waste items collected, ranging from years `r prof_trash_wheel |> pull(year) |> min()` to `r prof_trash_wheel |> pull(year) |> max()`. The key variables includes `dumpster`, `weight_tons`, `homes_powered`.

- For the `gwy_trash_wheel` dataset, there are `r nrow(gwy_trash_wheel)` observations and `r ncol(gwy_trash_wheel)` variables. This data also provides detail records of waste collection, capturing various information of waste items collected, ranging from years `r gwy_trash_wheel |> pull(year) |> min()` to `r gwy_trash_wheel |> pull(year) |> max()`. The key variables includes `dumpster`, `weight_tons`, `homes_powered`.

- In the resulting `trash_whell_combine` dataset, there are `r nrow(trash_wheel_combine)` observations and `r ncol(trash_wheel_combine)` variables. After cleaned and combined `mr_trash_wheel`, `prof_trash_wheel`, and `gwy_trash_wheel` by `year`, rows with missing dumpster values. Unnecessary columns `x15` and `x16` are dropped. `homes_powered` was calculated using the formula `(weight_tons * 500) / 30` and new variable name `wheel_name` column was created with the value "Mr. Trash Wheel", "Professor Trash Wheel" and "Gwynnda Trash Wheel" assigned to each dataset respectively for trash wheel identification purposes. 

- The total weight collected by Professor Trash Wheel was `r trash_wheel_combine |> filter(wheel_name == "Professor Trash Wheel") |> pull(weight_tons) |> sum()` tons

- There were `r trash_wheel_combine |> filter(wheel_name == "Gwynnda Trash Wheel", year == "2021", month == "July") |> pull(cigarette_butts) |> sum()` number of cigarette butts collected by Gwynnda in July of 2021


## Problem 3

### Clean `mci_baseline`

```{r clean_mci_baseline}
mci_baseline_full = 
  read_csv("data/data_mci/mci_baseline.csv", skip = 1, na = ".") %>% 
  janitor::clean_names() %>%
  mutate(
    sex = ifelse(sex == 1, "Male", "Female"),
    apoe4 = ifelse(apoe4 == 1, "Carrier", "Non carrier"),
    age_at_onset = as.numeric(age_at_onset)
  )

# The mci_baseline_full is the complete cleaned data but without dropping the participants who do not have mci

mci_baseline = 
  mci_baseline_full %>% 
  drop_na(age_at_onset) 

# mci_baseline is the final cleaned version after the participants don't have mci removed from mci_baseline_full
```

### Description and important values of`mci_baseline`

- In the `mci_baseline_full`, there are `r nrow(mci_baseline_full)` observations and `r ncol(mci_baseline_full)` variables. This data provided basic demographic information of participants in the study. 

- During the cleaning process, `janitor::clean_names()` function was used to standardize the variable names. `mutate` function changed the `sex` variable into `Male` and `Female` categories based on the original numeric coding of 1 and other values. Similarly, the `apoe4` variable is changed to `Carrier` and `Non carrier` categories. The `age_at_onset` is converted into numeric. Lastly, assigned a new dataset `mci_baseline`, which participants who do not have MCI excluded, and this is final version of baseline.

- There are a total of `r nrow(mci_baseline_full)` participants were recruited for the study. Of these participants, `r nrow(mci_baseline)` participants developed MCI. The average baseline age is `r mci_baseline |> pull(current_age) |> mean() |> round(2)`. Among the women in the study, `r round(nrow(filter(mci_baseline, sex == "Female", apoe4 == "Carrier"))/nrow(filter(mci_baseline, sex == "Female")) * 100, 2)`% are APOE4 carriers.



## Clean mci_amyloid

```{r clean_mci_amyloid, warning = FALSE}
mci_amyloid = 
  read_csv("data/data_mci/mci_amyloid.csv", skip = 1, na = ".") %>% 
  janitor::clean_names() %>% 
  rename(
    time_0 = baseline,
    id = study_id
  ) %>% 
  pivot_longer(
    time_0:time_8,
    names_to = "time_in_years",
    values_to = "amyloid_ratio"
  ) %>% 
  mutate(
    amyloid_ratio = as.numeric(amyloid_ratio)
  ) %>% 
  drop_na()



```

### Description of `mci_amyloid`

- In the `mci_amyloid`, there are `r nrow(mci_amyloid)` observations and `r ncol(mci_amyloid)` variables. This data provided information of participants in the study with `time_in_years` elapsed since the study baseline to the visit and also the biomarker `amyloid_ratio` was measure. 

- During the cleaning process, we skipped the first row and interpreting periods as missing value.  `janitor::clean_names()` function was used to standardize the variable names for consistency. The `rename` function is applied to change the variable name `baseline` to `time_0` for reading consistency; And `rename` function applied, so variable name `study_id` changed to `id` to match the `id` in `mci_baseline`. Next, the data transferred from wide format to long format by using the `pivot_longer` function. This process involves converting columns `time_0`, `time_2`, `time_4`, and `time_8` into two columns `time_in_years` and `amyloid_ratio` by using the `mutate` function. So, the `time_in_years` column now holds the time information, while `amyloid_ratio` column contains the corresponding `amyloid_ratio` values. Finally, `amyloid_ratio` is converted to numeric.


## Check the participant that only in `mci_baseline` or `mci_amyloid`

```{r}
mci_baseline_only = anti_join(mci_baseline, mci_amyloid, by = "id")
mci_amyloid_only = anti_join(mci_amyloid, mci_baseline, by = "id")
```

### Description of the `mci_baseline_only` and `mci_amyloid_only`

- After checking if there are participants appear only in `mci_baseline` or `mci_amyloid`, there are only `r nrow(mci_baseline_only)` participants appear in `mci_baseline` data, with `id`, `14`, `49`, `268`. On the other hand, there are `r nrow(mci_amyloid_only)` participants appear only in the `mci_amyloid` data.

## Combine both datasets 

```{r combine}
mci_combine = 
  inner_join(mci_baseline, mci_amyloid, by = "id") %>% 
  drop_na()

str(mci_combine)
```
### Description of the `mci_combine` data

- In the `mci_combine` dataset, the `inner_join` function was applied so that only participants who appear in both `mci_baseline` and `mci_amyloid` are retained.There are `r nrow(mci_combine)` observations and `r ncol(mci_combine)` variables. 

```{r csv_mci_combine}
write.csv(mci_combine, "data/mci_combine.csv")

```


